---
title: "Check-in 1: Instructions"
author: "Peter Gao"
date: "2023-08-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A. Running your first R code

1. On the computer you are using, create a new folder titled "MATH167R" or something similar.

2. Within this folder, create a sub-folder named "checkins" or something similar.

3. Download "[01_first-R-code.R](../static/01_first-R-code.R)" and save it in your checkins folder.

4. Open "01_first-R-code.R" in RStudio. It should look something like the code below.

```{r, file='../static/01_first-R-code.R', eval = F}
```

5. Run the code line by line and answer the two questions on the [Check-in 1 assignment](https://sjsu.instructure.com/courses/1571636/quizzes/1690553) on Canvas.


6. For complete credit for Check-in 1, please also complete the pre-course survey and NameCoach recording (links on Canvas).

7. If you would like to find others to work with, please also reply to the Piazza post.
