---
title: "Check-in 2"
author: "YOUR NAME HERE"
date: ""
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

Remember, **follow the instructions below and use R Markdown to create a pdf document with your code and answers to the following questions on Gradescope.** You may find a template file by clicking "Code" in the top right corner of this page.

### Collaborators

INSERT NAMES OF ANY COLLABORATORS

## A. Flights Data

**1.**  For this lab, we'll be using the `flights` data. You can load this data using the following code:

```{r}
library(nycflights13)
data(flights)
```

**2.** Use the `mutate()` function to turn `origin` into a factor.

```{r}
# YOUR CODE HERE
```

**3.** Compute the mean `arr_delay` for each origin airport. Which airport has the longest delays on average? You can drop missing rows.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->


**4.** Use the `mutate()` function to turn `month` into a factor.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->


**5.** Compute the mean `arr_delay` by month. What do you notice?

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->


**6.** Compute the mean `arr_delay` by month AND airport. What do you notice?

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->
