---
title: "Check-in 6"
author: "YOUR NAME HERE"
date: "2023-10-16"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit! In addition, to receive full credit, your code output and plots must be correctly formatted.

### Collaborators

INSERT NAMES OF ANY COLLABORATORS

Given a vector $\mathbf{x}=(x_1,\ldots, x_n)$ and a number $p\geq 1$, the vector p-norm is defined as

$$||\mathbf{x}||_p=(|x_1|^p+\cdots+|x_n|^p)^{1/p}$$

For example, the vector 2-norm is given by

$$||\mathbf{x}||_2=(|x_1|^2+\cdots+|x_n|^2)^{1/2}$$

Consider the following code.

```{r, eval = T, echo = T}
create_p_norm <- function(p = 2) {
  return(
    function(x) return(sum(abs(x) ^ p) ^ (1 / p))
  )
}
```

1. Explain in your own words what the following code does:

```{r, eval = T, echo = T}
norm_1 <- create_p_norm(1)
x <- c(-1, 1, 2)
norm_1(x)
```

2. Explain in your own words what the following code does:

```{r, eval = T, echo = T}
p <- 3
norm_mystery <- create_p_norm()
x <- c(-1, 1, 2)
norm_mystery(x)
```