---
title: "MATH167R: Tidying and manipulating data"
author: "Peter Gao"
format: 
  revealjs:
    theme: [./slides.scss, ../theme.scss]
editor: visual
---

## Overview of today

-   Tidying datasets
-   Exploratory analysis

## One dataset, four ways

::: columns
::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
library(tidyverse)
table1
```

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
table2
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
library(tidyverse)
table3
```

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
table4a
```

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
table4b
```
:::
:::

## Which way?

Suppose I wanted to create the following plot:

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
ggplot(table1, aes(x = year, y = cases)) +
  geom_line(aes(group = country), color = "grey50") +
  geom_point(aes(color = country, shape = country)) +
  scale_x_continuous(breaks = c(1999, 2000)) # x-axis breaks at 1999 and 2000
```

Which format should I use? Try to recreate this plot.

What if we wanted to compute the number of cases **per capita**, for each country and year?

## Munging data

What if we want to convert from `table2` or `table4a` and `table4b` to `table1`?

Today, we'll discuss how to do this in both **base R** and the **tidyverse**.

## `pivot_longer()` and `pivot_wider()`

The `pivot_longer()` and `pivot_wider()` functions are useful for reshaping data: `pivot_longer()` "lengthens" datasets and `pivot_wider()` does the opposite.

They are challenging to use because you need to have a clear vision of the data frame you want, not because they are complicated in R.

## Example: `pivot_longer()` {.smaller}

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
relig_income
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
library(knitr)
kable(relig_income) |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "480px")
```

## Example: `pivot_longer()` {.smaller}

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
relig_income |>
  pivot_longer(cols = !religion, names_to = "income", values_to = "count") |>
  ggplot(aes(x = income, fill = religion, y = count)) +
  geom_col() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.text = element_text(size = 7),
          legend.key.size = unit(.3, 'cm'))
```

## Example: `pivot_longer()`

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
#| code-line-numbers: "2"
relig_income |>
  pivot_longer(cols = !religion, names_to = "income", values_to = "count")
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
relig_income |>
  pivot_longer(cols = !religion, names_to = "income", values_to = "count") |>
  kable() |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "480px")
```

## Example: `pivot_longer()`

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
head(billboard)
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
kable(head(billboard)) |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "400px")
```

## Example: `pivot_longer()`

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    names_prefix = "wk",
    values_to = "rank",
    values_drop_na = TRUE
  ) |>
  head(n = 20) |>
  kable() |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "200px")
```

## Example: `pivot_wider()`

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
fish_encounters
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
kable(fish_encounters) |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "480px")
```

## Example: `pivot_wider()`

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
#| code-line-numbers: "2"
fish_encounters |>
  pivot_wider(names_from = station, values_from = seen)
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
fish_encounters |>
  pivot_wider(names_from = station, values_from = seen) |>
  head(n = 20) |>
  kable() |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "480px")
```

## Example: `pivot_wider()`

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
us_rent_income
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
kable(us_rent_income) |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "400px")
```

## Example: `pivot_wider()`

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
us_rent_income |>
  pivot_wider(
    names_from = variable,
    values_from = c(estimate, moe)
  )
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
us_rent_income |>
  pivot_wider(
    names_from = variable,
    values_from = c(estimate, moe)
  ) |>
  kable() |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "360px")
```

## One dataset, four ways

::: columns
::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
library(tidyverse)
table1
```

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
table2
```
:::

::: {.column width="50%"}
```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
library(tidyverse)
table3
```

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
table4a
```

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
table4b
```
:::
:::

## Munging data

What if we want to convert from `table2` or `table4a` and `table4b` to `table1`?

**Group work**: Work through how to convert `table2` to `table1` (and back). If time permits, try converting `table4a` and `table4b` `table1` (and back).

## Working with messier data: US population, 2010-2020

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
state_population <- readr::read_csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/national/totals/nst-est2020.csv")
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
# code to print the data
state_population <- readr::read_csv("https://www2.census.gov/programs-surveys/popest/datasets/2010-2020/national/totals/nst-est2020.csv")
library(knitr)
kable(state_population, digits = 3, row.names = F) |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "400px")
```

## National population data

We can use the `SUMLEV` variable to obtain only national data:

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
#| code-line-numbers: "6"
library(tidyverse)
US_pop <- state_population |>
  filter(SUMLEV == "010")
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
library(tidyverse)
US_pop <- state_population |>
  filter(SUMLEV == "010")
# code to print the data
kable(US_pop, digits = 3, row.names = F) |>
  kableExtra::kable_styling("striped", full_width = T)
```

## National population data

How can we generate the following plot?

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
US_pop |> 
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), names_to = "Year", values_to = "Population") |>
  mutate(Year = as.numeric(stringr::str_sub(Year, start = 12))) |>
  ggplot(aes(x = Year, y = Population)) + 
  geom_line()
```

## National population data

Note that our data cannot be readily used with `ggplot` to generate the previous plot. Ideally, we would like two vectors of the same length: one representing `Year` and one representing `Population`.

## Manipulating data for visualization

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
#| code-line-numbers: "6"
library(tidyverse)
US_pop <- state_population |>
  filter(SUMLEV == "010")
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
library(tidyverse)
US_pop <- state_population |>
  filter(SUMLEV == "010")
# code to print the data
kable(US_pop, digits = 3, row.names = F) |>
  kableExtra::kable_styling("striped", full_width = T)  |> 
  kableExtra::scroll_box(height = "120px")
```

## Approach 1: Build a data frame manually

```{r, echo = T, eval = F, error = T}
# base R
plot_dat <- data.frame(
  Year = 2010:2020,
  Population = US_pop[1, 8:19]
)
head(plot_dat)
```

. . .

```{r, echo = F, eval = T, error = T}
# base R
plot_dat <- data.frame(
  Year = 2010:2020,
  Population = US_pop[1, 8:19]
)
head(plot_dat)
```

## National population data

What went wrong? `US_pop[1, 8:19]` is a tibble, and we need a numeric.

. . .

```{r, echo = T, eval = F, error = T}
plot_dat <- data.frame(
  Year = 2010:2020,
  Population = as.numeric(US_pop[1, 8:19])
)
```

. . .

```{r, echo = F, eval = T, error = T}
# base R
plot_dat <- data.frame(
  Year = 2010:2020,
  Population = as.numeric(US_pop[1, 8:19])
)
```

## National population data

Now what? `US_pop` has an extra column `POPESTIMATE042020` which we need to remove.

. . .

```{r, echo = T, eval = T, error = T}
# base R
plot_dat <- data.frame(
  Year = 2010:2020,
  Population = as.numeric(US_pop[, c(8:17, 19)])
)
head(plot_dat)
```

## National population data

Finally, we can use the `plot` function to get our desired visualization:

```{r, echo = T, eval = T, error = T}
plot(plot_dat$Year, plot_dat$Population)
```

## Approach 2: Tidyverse

Now let's see how we might approach the same task in the tidyverse. Our strategy will be to "tidy" the data so we have one row for each observation and one column for each observation.

First, let's remove the unnecessary column using the `select()` function.

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "2"
US_pop |> 
  select(-POPESTIMATE042020)  
```

```{r}
#| echo: FALSE
#| eval: TRUE
#| code-line-numbers: "2"
US_pop |> 
  select(-POPESTIMATE042020)  |>
  kable() |> 
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "120px")
```

## Tidyverse strategy

Now let's use the `pivot_longer` function to combine the `POPESTIMATE` columns into two variables: `Year` and `Population`:

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "3-5"
US_pop |> 
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), 
               names_to = "Year",
               values_to = "Population")  
```

## Tidyverse strategy

```{r}
#| echo: FALSE
#| eval: TRUE
#| code-line-numbers: "2"
US_pop |> 
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), 
               names_to = "Year",
               values_to = "Population")  |>
  kable() |> 
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "400px")
```

What do we need to do next?

## Tidyverse strategy

Finally, we need to use the `str_sub()` function to shorten our `Year` variable before turning it into a numeric.

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "6"
US_pop |> 
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), 
               names_to = "Year",
               values_to = "Population") |>
  mutate(Year = as.numeric(stringr::str_sub(Year, start = 12)))
```

## Tidyverse strategy

Now, `Year` is numeric as desired:

```{r}
#| echo: FALSE
#| eval: TRUE
#| code-line-numbers: "2"
US_pop |> 
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), 
               names_to = "Year",
               values_to = "Population") |>
  mutate(Year = as.numeric(stringr::str_sub(Year, start = 12))) |>
  head(n = 30) |>
  kable() |> 
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "400px")
```

## Tidyverse strategy

Now we can `|>` it to `ggplot`:

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "7-8"
US_pop |> 
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), 
               names_to = "Year",
               values_to = "Population") |>
  mutate(Year = as.numeric(stringr::str_sub(Year, start = 12))) |>
  ggplot(aes(x = Year, y = Population)) + 
  geom_line()
```

## Tidyverse strategy

```{r}
#| echo: FALSE
#| eval: TRUE
#| code-line-numbers: "2"
#| fig-width: 7
#| fig-height: 4.75
US_pop |> 
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), 
               names_to = "Year",
               values_to = "Population") |>
  mutate(Year = as.numeric(stringr::str_sub(Year, start = 12))) |>
  ggplot(aes(x = Year, y = Population)) + 
  geom_line()
```

## Tidyverse strategy: More complicated plots

We can also work with the region level data, mapping the `color` aesthetic to the region:

```{r}
#| echo: TRUE
#| eval: FALSE
#| code-line-numbers: "2, 7-8"
state_population |> 
  filter(SUMLEV == "020") |>
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), 
               names_to = "Year",
               values_to = "Population") |>
  mutate(Year = as.numeric(stringr::str_sub(Year, start = 12))) |>
  ggplot(aes(x = Year, y = Population, color = NAME)) + 
  geom_line()
```

## Tidyverse strategy: More complicated plots

```{r}
#| echo: FALSE
#| eval: TRUE
#| code-line-numbers: "2, 7-8"
#| fig-width: 7
#| fig-height: 4.75
state_population |> 
  filter(SUMLEV == "020") |>
  select(-POPESTIMATE042020) |>
  pivot_longer(contains("POPESTIMATE"), 
               names_to = "Year",
               values_to = "Population") |>
  mutate(Year = as.numeric(stringr::str_sub(Year, start = 12))) |>
  ggplot(aes(x = Year, y = Population, color = NAME)) + 
  geom_line()
```

## Example: World Bank Inflation Data

The World Bank provides historical data on inflation in countries around the world [here](https://www.worldbank.org/en/research/brief/inflation-database).

In particular, let's explore the annual consumer price inflation data for the US.

. . .

The **consumer price inflation** measures the "annual percentage change in the cost to the average consumer of acquiring a basket of goods."

## Example: World Bank Inflation Data

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
library(openxlsx)
cpi <- read.xlsx("https://thedocs.worldbank.org/en/doc/1ad246272dbbc437c74323719506aa0c-0350012021/original/Inflation-data.xlsx",
                 sheet = 4)
cpi <- cpi[1:203,]
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
# code to print the data
library(openxlsx)
cpi <- read.xlsx("https://thedocs.worldbank.org/en/doc/1ad246272dbbc437c74323719506aa0c-0350012021/original/Inflation-data.xlsx",
                 sheet = 5)
cpi <- cpi[1:203,] # remove last row
kable(head(cpi, n = 25), digits = 3, row.names = F) |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "360px")
```

## Example: World Bank Inflation Data

How do we create a line plot with `Year` on the x-axis and `CPI` on the y-axis for the US?

. . .

```{r}
#| echo: true
#| eval: false
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
#| code-line-numbers: "2-4"
cpi |>
  pivot_longer(cols = `1970`:`2022`,
               names_to = "Year", 
               values_to = "CPI") |>
  mutate(Year = as.numeric(Year)) |>
  select(Country, Year, CPI)
```

```{r}
#| echo: false
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
cpi |>
  pivot_longer(cols = `1970`:`2022`,
               names_to = "Year", 
               values_to = "CPI") |>
  mutate(Year = as.numeric(Year)) |>
  select(Country, Year, CPI) |>
  head(n = 30) |>
  kable(digits = 3) |>
  kableExtra::kable_styling("striped", full_width = T) |> 
  kableExtra::scroll_box(height = "240px")
```

## Example: World Bank Inflation Data

```{r}
#| echo: true
#| eval: true
#| warning: false
#| fig-width: 7
#| fig-height: 4.75
#| out.width: "100%"
#| fig.align: "center"
cpi |>
  pivot_longer(cols = `1970`:`2022`,
               names_to = "Year", 
               values_to = "CPI") |>
  mutate(Year = as.numeric(Year)) |>
  filter(Country.Code == "USA") |>
  ggplot(mapping = aes(x = Year, y = CPI)) +
  geom_line()
```

## Data manipulation cheatsheet

-   `filter()` subset rows
-   `select()` subset columns, use with `contains()`, `starts_with()`, `ends_with()`, ...
-   `mutate()` create columns
-   `group_by()`, `summarize()`, `count()` group and summarize groups
-   `rename()` rename columns
-   `pivot_longer()`, `pivot_wider()` reshape data
-   `inner_join()`, `left_join()`, `right_join()`, `outer_join()` combine data (like SQL)

[Much, much more! (Click me for a cheat sheet)](https://rstudio.com/resources/cheatsheets/)

## Exploratory data analysis

Exploratory data analysis is the process of learning about a dataset by visualizing, summarizing, and transforming it.

By now we've seen most of the skills you need to carry out exploratory data analysis (EDA). As Wickham, Çetinkaya-Rundel, and Grolemund put it, the cycle of EDA looks something like this:

1.  Generate questions about your data.

2.  Explore your questions by visualizing, transforming, and modelling your data.

3.  Use your findings to refine your questions and generate new questions.

You can read their walkthrough of the EDA process [here](https://r4ds.hadley.nz/eda).
