---
title: "MATH167R: Simulations"
author: "Peter Gao"
format: 
  revealjs:
    theme: [./slides.scss, ../theme.scss]
editor: visual
---

## Overview of today

-   Random variables in R
-   The `r`, `p`, `d`, and `q` functions
-   `replicate()`
-   Writing simulations

## Random variables in R

We have already seen a number of functions for generating random variables such as `sample()` and `rnorm()`.

Today, we will expand on these functions and related functions for studying random variables and begin using them to write various simulations.

## The `r`, `p`, `d`, and `q` functions

1.  `rnorm()` can be used to generate `n` random observations from a normal distribution:

```{r}
#| echo: TRUE
#| eval: TRUE
rnorm(n = 1, mean = 0, sd = 1)
```

2.  `pnorm()` can be used to compute the distribution function at a value `q`:

```{r}
#| echo: TRUE
#| eval: TRUE
pnorm(q = 1, mean = 0, sd = 1)
```

3.  `qnorm()` can be used to compute the quantile function at a percentile `p`:

```{r}
#| echo: TRUE
#| eval: TRUE
qnorm(p = .75, mean = 0, sd = 1)
```

4.  `dnorm()` can be used to compute the density function at a value `x`:

```{r}
#| echo: TRUE
#| eval: TRUE
dnorm(x = .5, mean = 0, sd = 1)
```

## Exercise

Suppose we perform a two-sided hypothesis test using a normal distribution and obtain a z-score of 1.24. How would we compute the p-value?

. . .

```{r}
#| echo: TRUE
#| eval: TRUE
2 * (1 - pnorm(q = 1.24, mean = 0, sd = 1))
```

. . .

Suppose we perform a two-sided hypothesis test using a normal distribution and obtain a z-score of -2.12. How would we compute the p-value?

. . .

```{r}
#| echo: TRUE
#| eval: TRUE
2 * pnorm(q = 2.12, mean = 0, sd = 1)
```

## Random seeds

Remember that `set.seed()` can be used to ensure that you obtain the same results each time you run your code.

For example, if you include the command `x <- rnorm(1)` in an .Rmd document with running `set.seed()`, each time you knit, you will produce a different value of `x`.

```{r}
#| echo: TRUE
#| eval: TRUE
set.seed(1022)
rnorm(n = 1, mean = 0, sd = 1)
```

## Monte Carlo methods

Monte Carlo methods or Monte Carlo experiments use repeated random sampling to obtain numerical results or to approximate quantities of interest.

Monte Carlo experiments are often used to:

-   approximate integrals
-   generate draws from a probability distribution
-   optimize functions stochastically

## Example: Transformations of random variables

Suppose $X\sim N(0, 1)$. What is $E(e^X)$?

We could use the change of variable formula to compute the expectation. Or we could use simulation to approximate this quantity. Asymptotic analysis is needed to study the convergence of our approximation.

```{r}
#| echo: TRUE
#| eval: TRUE
#| error: TRUE
n_sim <- 1000
n <- 100
exp_X <- numeric(n_sim)
for (i in 1:n_sim) {
  exp_X[i] <- exp(rnorm(n))
}
mean(exp_X)
```

## Example: Maximum of random variables

Suppose $X_i\sim \mathrm{Exponential}(1)$ for $1\leq i\leq n$, for some $n$.

-   What is $E(\mathrm{max}(X_1,\ldots, X_n))?$
-   How does $E(\mathrm{max}(X_1,\ldots, X_n))$ change for different values of $n$?

```{r}
#| echo: TRUE
#| eval: TRUE
set.seed(1022)
n_sim <- 1000
n <- 10
maxes <- numeric(n_sim)
for (i in 1:n_sim) {
  maxes[i] <- max(rexp(n, rate = 1))
}
mean(maxes)
```

. . .

What about for $n=100$? Should the expected maximum be larger or smaller?

## Example: Maximum of random variables

```{r}
#| echo: TRUE
#| eval: TRUE
n_sim <- 1000
n <- 100
maxes <- numeric(n_sim)
for (i in 1:n_sim) {
  maxes[i] <- max(rexp(n, rate = 1))
}
mean(maxes)
```

## `replicate()` for repeated evaluation

In the previous code, we have been using `for` loops, which explicitly and repeatedly change global variables.

We can alternatively use the `replicate()` function to repeatedly evaluate an expression and in particular to repeatedly generate data. The data will be automatically organized into a matrix or vector.

## `replicate()` for repeated evaluation

For example, we can generate `n` exponential random variables and take their maximum:

```{r}
#| echo: TRUE
#| eval: TRUE
replicate(10, rexp(10)) # matrix output
replicate(10, max(rexp(10))) # vector output
```

## `replicate()` for repeated evaluation

Using `replicate()`, the previous simulation is just:

```{r}
#| echo: TRUE
#| eval: TRUE
maxes <- replicate(1000, max(rexp(n)))
mean(maxes)
```

## Example: Maximum of random variables

We can repeat this for many different values of `n`.

```{r declarative.sim}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
n_sim <- 1000
n_values <- c(10, 25, 50, 100, 250, 500, 1000)
max_by_n <- c()
for (n in n_values) {
  maxes <- numeric(n_sim)
  for (i in 1:n_sim) {
    maxes[i] <- max(rexp(n, rate = 1))
  }
  max_by_n <- c(max_by_n, mean(maxes))
}
max_by_n
```

## Example: Maximum of random variables

Note that we can write the maximum of exponentials simulation using `replicate()`:

```{r functional.sim}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
max_by_n <- vapply(
  n_values, 
  function(n) mean(replicate(n_sim, max(rexp(n, rate = 1)))), 
  numeric(1)
)
max_by_n
```

## Example: Maximum of random variables

It turns out that $\mathrm{max}(X_1,\ldots, X_n)$ can be modeled using a Gumbel distribution (after being transformed). Below, we plot the empirical mean and approximate the theoretical expectation (for large $n$).

```{r plot_results}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
plot(n_values, max_by_n,
     type = "l",
     xlab = "n", ylab = "Maximum",
     main = paste0("Sample mean of maximums (black)",
                   "and theoretical expectation (red)"))
lines(n_values, -digamma(1) + log(n_values), col = "red")
```

## Random walks

Random walks are stochastic processes that describe a sequence of random steps on some space.

A simple example is the one-dimensional random walk on the integer number line which starts at 0 and for each step, moves forward (+1) or backward (-1) with equal probability.

## Random walks

We can easily simulate a one-dimensional random walk:

```{r random_walk}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
set.seed(123)
n_steps <- 100
x <- numeric(n_steps)
x[1] <- 0
for (i in 2:n_steps) {
  x[i] <- x[i - 1] + sample(c(1, -1), 1)
}
```

. . . We can then plot our results

```{r random_walk_plot}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
plot(1:n_steps, x, type = "l",
     xlab = "Step", ylab = "x",
     main = "A one-dimensional random walk")
```

## A more functional version with recursion

We can also adopt a more functional approach:

```{r random_walk_functional}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
take_step <- function() {
  return(sample(c(1, -1), 1))
}
walk_randomly <- function(n_steps, start = 0) {
  if (n_steps <= 1) {
    return(start)
  }
  x <- c(start, walk_randomly(n_steps - 1, start + take_step()))
  return(x)
}
plot(1:100, walk_randomly(100), type = "l",
     xlab = "Step", ylab = "x",
     main = "A one-dimensional random walk")
```

## A two-dimensional random walk

A two-dimensional random walk takes place on the integer grid, where at each step, we can either walk north, south, east, or west.

```{r 2d_random_walk_functional}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
take_step_2d <- function() {
  steps <- rbind(c(0, 1),
                 c(0, -1), 
                 c(1, 0),
                 c(-1, 0))
  return(steps[sample(1:4, 1), ])
}
walk_randomly_2d <- function(n_steps, start = c(0, 0)) {
  if (n_steps <= 1) {
    return(start)
  }
  x <- rbind(start, walk_randomly_2d(n_steps - 1, start + take_step_2d()))
  rownames(x) <- NULL
  return(x)
}
n_steps <- 100
coords <- walk_randomly_2d(n_steps)
coords
```

## A two-dimensional random walk

We can use `geom_path()` to illustrate the path taken by the random walker:

```{r 2d_random_walk_functional_plot}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
library(ggplot2)
coords_dat <- data.frame(x = coords[, 1],
                         y = coords[, 2],
                         step = 1:n_steps)
ggplot(coords_dat, aes(x = x, y = y, color = step)) + geom_path()
```

```{r}
knitr::knit_exit()
```

## A two-dimensional random walk

We can use the `gganimate` package to animate our visualization:

```{r 2d_random_walk_functional_anim}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
library(gganimate)
anim_walk <- ggplot(coords_dat, aes(x = x, y = y, color = step)) +
  geom_path() +
  transition_reveal(along = step)
animate(anim_walk)
```
