---
title: "MATH167R: Sampling distributions"
author: "Peter Gao"
format: 
  revealjs:
    theme: [./slides.scss, ../theme.scss]
editor: visual
---

## Overview of today

-   Sampling distributions
-   Bootstrap distributions

## Sampling distributions

Without explicitly saying so, we have been simulating sampling distributions during our simulations.

Recall: A sampling distribution is the probability distribution of a sample-based statistic.

Example: If $X_1,\ldots, X_{100}$ are iid random variables with variance $1$, what is the distribution of the sample mean $\overline{X}$?

## Simulating a sampling distribution

Example simulation: Suppose $X_1,\ldots, X_{100}\sim N(0,1)$.

```{r sample_mean}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
set.seed(123)
Xbar <- replicate(10000, mean(rnorm(100)))
hist(Xbar, main = "Sampling distribution of mean of 100 N(0, 1) variables")
```

## Simulating a sampling distribution

What if $X_1,\ldots, X_{100}\sim \mathrm{Exp}(1)$?

```{r sample_mean_exp}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
set.seed(123)
Xbar <- replicate(10000, mean(rexp(100, rate = 1)))
hist(Xbar, main = "Sampling distribution of mean of 100 Exponential(1) variables")
```

## Simulating a sampling distribution

What if $X_1,\ldots, X_{100}\sim \mathrm{Poisson}(1)$?

```{r sample_mean_pois}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
set.seed(123)
Xbar <- replicate(10000, mean(rpois(100, lambda = 1)))
hist(Xbar, main = "Sampling distribution of mean of 100 Poisson(1) variables")
```

## Simulating a sampling distribution

By the central limit theorem, all of these sampling distributions are asymptotically Gaussian with variance 1 (though the mean values differ between the three). Since $n=100$ is fairly large (for these examples), all of the previous examples are close to Gaussian.

What about if $n$ is smaller? Or if we consider other distributions?

Simulation is a powerful strategy for testing out methods:

-   Suppose you wish to know... is $n$ large?
-   Create a simulation that generates data that looks like your real world data.
-   Evaluate whether the results are reasonable/accurate using simulated data.

## Smaller sample sizes

```{r smaller_sample}
#| echo: TRUE
#| eval: FALSE
#| cache: TRUE
set.seed(123)
Xbar_normal <- replicate(10000, mean(rnorm(10)))
Xbar_exp <- replicate(10000, mean(rexp(10, rate = 1)))
Xbar_poisson <- replicate(10000, mean(rpois(10, lambda = 1)))
par(mfrow = c(1, 3))
hist(Xbar_normal, main = "N(0, 1)")
hist(Xbar_exp, main = "Exp(1)")
hist(Xbar_poisson, main = "Poisson(1)")
```


## Smaller sample sizes

```{r smaller_sample2}
#| echo: FALSE
#| eval: TRUE
#| cache: TRUE
set.seed(123)
Xbar_normal <- replicate(10000, mean(rnorm(10)))
Xbar_exp <- replicate(10000, mean(rexp(10, rate = 1)))
Xbar_poisson <- replicate(10000, mean(rpois(10, lambda = 1)))
par(mfrow = c(1, 3))
hist(Xbar_normal, main = "N(0, 1)")
hist(Xbar_exp, main = "Exp(1)")
hist(Xbar_poisson, main = "Poisson(1)")
```

## Other statistics

What if we want the sampling distribution of the maximum? For each of these population models, the sampling distribution of the maximum is different.

```{r max_sampling_dist}
#| echo: TRUE
#| eval: FALSE
#| cache: TRUE
set.seed(123)
Xmax_normal <- replicate(10000, max(rnorm(10)))
Xmax_exp <- replicate(10000, max(rexp(10, rate = 1)))
Xmax_poisson <- replicate(10000, max(rpois(10, lambda = 1)))
par(mfrow = c(1, 3))
hist(Xmax_normal, main = "N(0, 1)")
hist(Xmax_exp, main = "Exp(1)")
hist(Xmax_poisson, main = "Poisson(1)")
```

## Other statistics

What if we want the sampling distribution of the maximum? For each of these population models, the sampling distribution of the maximum is different.

```{r max_sampling_dist2}
#| echo: FALSE
#| eval: TRUE
#| cache: TRUE
set.seed(123)
Xmax_normal <- replicate(10000, max(rnorm(10)))
Xmax_exp <- replicate(10000, max(rexp(10, rate = 1)))
Xmax_poisson <- replicate(10000, max(rpois(10, lambda = 1)))
par(mfrow = c(1, 3))
hist(Xmax_normal, main = "N(0, 1)")
hist(Xmax_exp, main = "Exp(1)")
hist(Xmax_poisson, main = "Poisson(1)")
```

## Transformations

We can easily simulate sampling distributions for statistics summarizing the distribution of random variables that are defined as transformations of other random variables.

```{r mean_transformed}
#| echo: TRUE
#| eval: FALSE
#| cache: TRUE
set.seed(123)
Xbar_normal <- replicate(10000, mean(rnorm(10) ^ 2))
Xbar_exp <- replicate(10000, mean(rexp(10, rate = 1) ^ 2))
Xbar_poisson <- replicate(10000, mean(rpois(10, lambda = 1) ^ 2))
par(mfrow = c(1, 3))
hist(Xbar_normal, main = "N(0, 1) squared")
hist(Xbar_exp, main = "Exp(1) squared")
hist(Xbar_poisson, main = "Poisson(1) squared")
```

## Transformations

We can easily simulate sampling distributions for statistics summarizing the distribution of random variables that are defined as transformations of other random variables.

```{r mean_transformed2}
#| echo: FALSE
#| eval: TRUE
#| cache: TRUE
set.seed(123)
Xbar_normal <- replicate(10000, mean(rnorm(10) ^ 2))
Xbar_exp <- replicate(10000, mean(rexp(10, rate = 1) ^ 2))
Xbar_poisson <- replicate(10000, mean(rpois(10, lambda = 1) ^ 2))
par(mfrow = c(1, 3))
hist(Xbar_normal, main = "N(0, 1) squared")
hist(Xbar_exp, main = "Exp(1) squared")
hist(Xbar_poisson, main = "Poisson(1) squared")
```

## Exercise

-   Write code to create a histogram of the sampling distribution of the sample mean $\overline{Y}$ of $Y_1,\ldots, Y_{10}$ where $Y_i=\sin(X_i)$ and $X_1,\ldots, X_{10}$ are iid Uniform(0, 1) random variables.

## Resampling methods

The previous simulations are all based on parametric models (normal, exponential, Poisson). However, we can also simulate sampling distributions without parametric assumptions on the data generating mechanism.

## Resampling methods

Suppose we know $X_1,\ldots, X_{20}\sim \mathrm{Exp}(1)$. The sampling distribution of the sample mean can be simulated as follows:

```{r exp_samp_dist}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
set.seed(123)
Xbar_exp <- replicate(10000, mean(rexp(20, rate = 1)))
hist(Xbar_exp)
```

## Resampling methods

What if we observe $X_1,\ldots, X_{20}$, but we don't know that $X_i$ is exponentially distributed? We could just try the normal approximation to approximate the sampling distribution.

```{r exp_samp_dist_2, fig.height = 4, fig.width = 5}
#| echo: TRUE
#| eval: FALSE
#| cache: TRUE
X <- rexp(20, rate = 1)
x_axis <- seq(0, 3, length.out = 1000)

# plot "true" simulated sampling distribution
hist(Xbar_exp, freq = F, ylim = c(0, 3))

# plot normal theory
lines(x_axis, dnorm(x_axis, mean = mean(X), sd = sd(X) / sqrt(20)), col = 'red')
```

## Resampling methods

```{r exp_samp_dist_22, fig.height = 4, fig.width = 5}
#| echo: FALSE
#| eval: TRUE
#| cache: TRUE
X <- rexp(20, rate = 1)
x_axis <- seq(0, 3, length.out = 1000)

# plot "true" simulated sampling distribution
hist(Xbar_exp, freq = F, ylim = c(0, 3))

# plot normal theory
lines(x_axis, dnorm(x_axis, mean = 1, sd = 1 / sqrt(20)), col = 'red')
```

## Bootstrap samples

Bootstrapping is a resampling method that uses random samples with replacement from the sample data to approximate the sampling distribution of sample-based statistics.

```{r exp_samp_dist_3}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
print(sort(X))
# sample with replacement from X
X_resample <- sample(X, size = 20, replace = TRUE)
print(sort(X_resample))
```

## Bootstrap samples

We can repeatedly resample from $X$, then compute the resample sample means and plot the distribution:

```{r exp_samp_dist_4}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
bootstrap_Xbar <- replicate(10000, mean(sample(X, size = 20, replace = TRUE)))
hist(bootstrap_Xbar - mean(X) + 1)
```

## Bootstrap samples

We can compare the bootstrap sampling distributions and true sampling distributions for $\overline{X}-E(\overline{X})$.

```{r exp_samp_dist_5}
#| echo: TRUE
#| eval: FALSE
#| cache: TRUE
library(ggplot2)
plot_dat <-
  data.frame(Xbar = c(Xbar_exp - 1, bootstrap_Xbar - mean(X)),
             source = rep(c("Parametric (Exp)", "Bootstrap"), each = 10000))
ggplot(plot_dat, aes(x = Xbar, color = source)) +
  geom_density() +
  geom_function(fun = dnorm, 
                args = list(mean = 0, sd = 1 / sqrt(20)),
                aes(color = "Parametric (Normal)")) +
  xlab("Xbar - E(Xbar)")
```

## Bootstrap samples

We can compare the bootstrap sampling distributions and true sampling distributions for $\overline{X}-E(\overline{X})$.

```{r exp_samp_dist_52}
#| echo: FALSE
#| eval: TRUE
#| cache: TRUE
library(ggplot2)
plot_dat <-
  data.frame(Xbar = c(Xbar_exp - 1, bootstrap_Xbar - mean(X)),
             source = rep(c("Parametric (Exp)", "Bootstrap"), each = 10000))
ggplot(plot_dat, aes(x = Xbar, color = source)) +
  geom_density() +
  geom_function(fun = dnorm, 
                args = list(mean = 0, sd = 1 / sqrt(20)),
                aes(color = "Parametric (Normal)")) +
  xlab("Xbar - E(Xbar)")
```

## Bootstrap confidence intervals

We can take appropriate quantiles of the bootstrap samples to construct a bootstrap confidence interval using the `quantile()` function:

```{r exp_samp_dist_6}
#| echo: TRUE
#| eval: TRUE
#| cache: TRUE
quantile(bootstrap_Xbar, probs = c(.025, .975))
```

## Why bootstrap?

In this case, inference based on bootstrapping does not seem better than the normal approximation based on the central limit theorem.

However, for many cases, we may not have a simple parametric approximation to the true sampling distribution. For example, suppose we wish to compute the sampling distribution of the sample median.

Note however that the bootstrap may not work well for all statistics, such as the minimum (why not?).
