---
title: "Lab 2"
author: "YOUR NAME HERE"
date: "2023-08-30"
output: html_document
---

You should edit this .Rmd using RStudio, then click *Render* in the menu bar of the Source window (above the text of this .Rmd). Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit!

### Collaborators

INSERT NAMES OF ANY COLLABORATORS

## A. Functions and Vectors

In this problem, we will revisit the `sample` function to simulate the birthday problem. The birthday problem asks for the probability that in a set of $n$ randomly chosen people, at least two will share a birthday. For the purpose of this problem, suppose that we live in a universe where there are exactly 365 days in a year (sorry to those of you with leap day birthdays!) and each individual has an equal probability of being born on any given day.

**1.** Write code to create a vector `birthdays` of the numbers 1:365.

    ```{r, eval = T}
    # YOUR CODE HERE
    ```

**2.** Use the `sample()` function to sample $n=5$ birthdays (with replacement) from `birthdays` and save it as a vector `sample_5`. Repeat this with $n=10, 20, 30, 50, 100$. You may want to use `set.seed()` to set a random seed.

    ```{r, eval = T}
    # YOUR CODE HERE
    ```

**3.** Did you sample the same birthday twice in `sample_5`? What about for the other samples? Write code that can be used to test whether any of the days appears twice in each of your samples. There are many ways to do this and you can use any functions. In particular, `length()` and `unique()` may be useful.

    ```{r, eval = T}
    # YOUR CODE HERE
    ```

**4.** What do you observe? Discuss any patterns or surprising findings.

<!-- REPLACE THIS LINE WITH YOUR ANSWER -->

## B. Working with Data Frames

Use the following code to load the `penguins` data.

```{r, eval = T}
# load palmer penguins package
library(palmerpenguins)

# open penguins data as a data frame
data(penguins)
penguins <- as.data.frame(penguins)
```

**5.** Using the `mean()` function, compute the mean body mass of penguins in the dataset, dropping any missing values.

    ```{r, eval = T}
    # YOUR CODE HERE
    ```
    
    <!-- REPLACE THIS LINE WITH YOUR ANSWER -->
    
**6.** Using the `max` function, compute the maximum flipper length of penguins in the dataset, dropping any missing values.

    ```{r, eval = T}
    # YOUR CODE HERE
    ```
    
    <!-- REPLACE THIS LINE WITH YOUR ANSWER -->
    
**7.** Using the `hist` function, create a histogram of the ratio of the penguins' bill length to bill depth, dropping any missing values. What is the shape of the reuslting distribution?

    ```{r, eval = T}
    # YOUR CODE HERE
    ```
    
    <!-- REPLACE THIS LINE WITH YOUR ANSWER -->
