---
title: "Lab 3"
author: "YOUR NAME HERE"
date: "2023-09-06"
output: html_document
---

You should edit this .Rmd using RStudio, then click *Render* in the menu bar of the Source window (above the text of this .Rmd). Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit!

### Collaborators

INSERT NAMES OF ANY COLLABORATORS

```{r}
# LOAD ANY RELEVANT PACKAGES HERE
```

## A. Weather Forecast Data

1.  For this lab, we'll be using data on weather forecasts gathered by student at Saint Louis University. You can read about the dataset [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-12-20). Download the weather forecasts data using the following code:

    ```{r}
    weather_forecasts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-12-20/weather_forecasts.csv')
    ```

2.  How many rows are in this dataset? How many columns?

    ```{r}
    # YOUR CODE HERE
    ```

    <!--# YOUR ANSWER/EXPLANATION HERE -->

3.  How many cities are represented in this dataset?

    ```{r}
    # YOUR CODE HERE
    ```

    <!--# YOUR ANSWER/EXPLANATION HERE -->
    
4.  Create a new data frame containing only the forecasts for San Jose. You may have to explore the values for the `city` variable.

    ```{r}
    # YOUR CODE HERE
    ```

    <!--# YOUR ANSWER/EXPLANATION HERE -->

5. Compute the mean absolute error between `observed_temp` and `forecast_temp` for San Jose.

    ```{r}
    # YOUR CODE HERE
    ```

    <!--# YOUR ANSWER/EXPLANATION HERE -->
    
6. Compute the mean absolute error between `observed_temp` and `forecast_temp` for San Jose using only forecasts made 48 hours in advance.

    ```{r}
    # YOUR CODE HERE
    ```

    <!--# YOUR ANSWER/EXPLANATION HERE -->

7. Compute the mean absolute error between `observed_temp` and `forecast_temp` for San Jose using only forecasts made 12 hours in advance.

    ```{r}
    # YOUR CODE HERE
    ```

    <!--# YOUR ANSWER/EXPLANATION HERE -->

8. Compare your answers to 6 and 7. What do you notice? 

    <!--# YOUR ANSWER/EXPLANATION HERE -->




