---
title: "Lab 3: Descriptive Statistics"
author: "YOUR NAME HERE"
date: 2024-02-14
date-format: "[Due] MMMM DD, YYYY"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"

---

You should edit this .Rmd using RStudio, then click *Render* in the menu bar of the Source window (above the text of this .Rmd). Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit!

### Collaborators

INSERT NAMES OF ANY COLLABORATORS

```{r}
# LOAD ANY RELEVANT PACKAGES HERE
```

## A. Weather Forecast Data

1.  For this lab, we'll be using data on weather forecasts gathered by student at Saint Louis University. You can read about the dataset [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2022/2022-12-20). Download the weather forecasts data using the following code:

```{r}
weather_forecasts <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-12-20/weather_forecasts.csv')
```

2.  How many rows are in this dataset? How many columns?

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

3.  How many cities are represented in this dataset?

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

4.  Create a new data frame containing only the forecasts for San Jose. You may have to explore the values for the `city` variable.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

5. Compute the mean absolute error between `observed_temp` and `forecast_temp` for San Jose.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

6. Compute the mean absolute error between `observed_temp` and `forecast_temp` for San Jose using only forecasts made 48 hours in advance.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

7. Compute the mean absolute error between `observed_temp` and `forecast_temp` for San Jose using only forecasts made 12 hours in advance.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

8. Compare your answers to 6 and 7. What do you notice? How does this compare to your expectation?

<!--# YOUR ANSWER/EXPLANATION HERE -->

9. Pick two cities in this dataset. Investigate whether the forecast accuracy is better for one city than for the other, using an appropriate statistic. Discuss your findings.

<!--# YOUR ANSWER/EXPLANATION HERE -->

## B. Find your own data

For this component, pick a [Tidy Tuesday dataset](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023) and complete the following activity.

10. Provide a brief description of your dataset. Identify one question you could try to answer using this dataset.

<!--# YOUR ANSWER/EXPLANATION HERE -->

11. Open your dataset in R and compute one or more descriptive statistics that shed light on your question. Discuss your findings.

<!--# YOUR ANSWER/EXPLANATION HERE -->

12. Are there any limitations of your analysis? Could additional data or more complicated methods improve your analysis? Discuss.

<!--# YOUR ANSWER/EXPLANATION HERE -->
