---
title: "Lab 4: Data Visualization"
author: "YOUR NAME HERE"
date: "2023-09-13"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit! In addition, to receive full credit, your code output and plots must be correctly formatted.

### Collaborators

INSERT NAMES OF ANY COLLABORATORS

```{r}
# LOAD ANY RELEVANT PACKAGES HERE
```

## A. Basic visualizations

For this portion, we'll be using the `palmerpenguins` data. Use the following code to load the data.

```{r}
library(palmerpenguins)
data(penguins)
```

1.  Create and interpret a histogram of `bill_length_mm` using base R code. Be sure to use meaningful axis labels and titles.

2.  Create and interpret a histogram of `bill_length_mm` using ggplot2. Be sure to use meaningful axis labels and titles.

3.  Create and interpret a scatterplot of `bill_length_mm` versus `bill_depth_mm` using base R code. Be sure to use meaningful axis labels and titles.

4. Create and interpret a scatterplot of `bill_length_mm` versus `bill_depth_mm` using ggplot2. Be sure to use meaningful axis labels and titles.

5. Update your `ggplot2` scatterplot of `bill_length_mm` versus `bill_depth_mm` using ggplot2 so that the color of a point represents the corresponding penguin's species. What do you notice?


## B. Analyzing trends in San Jose rental prices 

For this component, you will be exploring and visualizing data on Craigslist apartment rental postings in the Bay Area. The data are available [here](https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-07-05/rent.csv) from Tidy Tuesday, as prepared by [Dr. Kate Pennington](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-07-05/readme.md). Note that you can use links within `read_csv()` to read online .csv files. I recommend saving a version of the unprocessed .csv on your machine in a `data` subfolder within your project folder so you will be able to work offline.

6. How many 1 bedroom listings from Santa Clara county are in this dataset?

7. What is the median price for a 1 bedroom listing in Santa Clara county in 2018?

8. Which county has the highest median price for a 1 bedroom listing in 2018?

9. Create two histograms for the prices of 1 bedroom listings in Santa Clara county in 2005 and 2018. Compare and discuss.

10. Create and interpret a line plot with year on the x-axis and median price for a 1 bedroom apartment for Santa Clara county on the y-axis from 2000 to 2018. 

11. Create and interpret a single plot with year on the x-axis and median price for a 1 bedroom apartment on the y-axis, using a separate line for each city in Santa Clara county, for the years 2000 to 2018. 

## C. Open ended data visualization

For this part, choose a dataset that interests you and identify a set of questions that you would like to explore via data visualizations. In particular, you should create three visualizations that satisfy the following requirements.

### Instructions

* Identify three research questions of interest that you want to study using this dataset.
* For each of your three research questions, generate a data visualization using your dataset. Discuss and interpret your findings. 
* Your project should include at least two different types of visualizations (e.g. scatterplots, box plots, bar plots, histograms, line plots, etc.).
* At least one of your plots should display variation over time or location (or both) in some way.
* Each visualization should include a caption that fully explains how to understand your visualization (i.e. explain all the components, legends, etc.). A good guideline is that someone who has not read your report should be able to look at just a visualization and its caption and fully understand what that visualization is showing.
* Each visualization must be accompanied by at least one paragraph of text. This text should include an interpretation of your visualization as well as what is interesting about your visualization. A strong visualization will be accompanied by text explaining what patterns or insights it helps us glean from the data. 
