---
title: "Lab 4"
author: "YOUR NAME HERE"
date: "2023-09-13"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit! In addition, to receive full credit, your code output and plots must be correctly formatted.

### Collaborators

INSERT NAMES OF ANY COLLABORATORS

```{r}
# LOAD ANY RELEVANT PACKAGES HERE
```

## A. Penguins visualizations

1.  For this lab, we'll be using the `palmerpenguins` data. Use the following code to load the data.

```{r}
library(palmerpenguins)
data(penguins)
```

2.  Create and interpret a histogram of `bill_length_mm` using base R code. Be sure to use meaningful axis labels and titles.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

3.  Create and interpret a histogram of `bill_length_mm` using ggplot2. Be sure to use meaningful axis labels and titles.

```{r}
# YOUR CODE HERE
```

4.  Create and interpret a scatterplot of `bill_length_mm` versus `bill_depth_mm` using base R code. Be sure to use meaningful axis labels and titles.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

5. Create and interpret a scatterplot of `bill_length_mm` versus `bill_depth_mm` using ggplot2. Be sure to use meaningful axis labels and titles.

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

6. Update your `ggplot2` scatterplot of `bill_length_mm` versus `bill_depth_mm` using ggplot2 so that the color of a point represents the corresponding penguin's species. What do you notice?

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

## B. Your choice of data

7. Find and download a dataset from the Tidy Tuesday project (https://github.com/rfordatascience/tidytuesday). 

```{r}
# YOUR CODE HERE
```

8. Create and interpret a univariate visualization for a variable of your choice. 

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->

9. Create and interpret a multivariate visualization for a variable of your choice. 

```{r}
# YOUR CODE HERE
```

<!--# YOUR ANSWER/EXPLANATION HERE -->


