---
title: "Lab 5"
author: "YOUR NAME HERE"
date: "2023-09-20"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit! In addition, to receive full credit, your code output and plots must be correctly formatted.

### Collaborators

For this lab, please work with one to two other students. List their names here.

```{r}
# LOAD ANY RELEVANT PACKAGES HERE
```

## A. 

1. Download the median age by state tables for 2010 and 2020 from the Canvas assignment [here](https://sjsu.instructure.com/courses/1571636/assignments/6775084). You can use the `read_csv` function from the `readr` package to open these files and save them as `age_2010` and `age_2020`.
 
```{r, eval = F, echo = T}

```

2. Create a scatter plot with each state as a point, with 2020 median age on the y-axis and 2010 median age on the x-axis. You may use any of the functions we have discussed in class. Note that there are many possible solutions. My solution used `pivot_wider()`, `left_join()`, and `ggplot()`, but you are welcome to use any solution you wish.


**Note**: the following commands were used to download this data using the `tidycensus` package. You do not need to run the following code: it is only provided in case you are curious about downloading data from the U.S. Census using R. In your final report, be sure that the following code chunk is set to `eval = F` and `echo = F`.

```{r, eval = F, echo = T}
library(tidycensus)

# the following code loads my local Census API Key, which is used to access 
# census data
Sys.getenv("CENSUS_API_KEY")

# 2010 median age by state
age_2010 <- get_decennial(geography = "state", 
                          variables = "P013I001", 
                          year = 2010,
                          sumfile = "sf1")

# 2020 median age by state
age_2020 <- get_decennial(geography = "state",
                          variables = "P13_001N",
                          year = 2020,
                          sumfile = "dhc")


# write to csv files
write.csv(age_2010, "~/Documents/MATH167R/data/median_age_by_state_2010.csv")
write.csv(age_2020, "~/Documents/MATH167R/data/median_age_by_state_2020.csv")
```

## B. Your choice of data

3. As a group, identify a dataset of interest and open it in R. Provide a brief one paragraph description of the dataset and explain why you chose it.

4. State three possible research questions you could explore using this dataset.

5. For each of your three research questions, generate a descriptive statistic or data visualization using your dataset. Discuss and interpret your findings.

