---
title: "Lab 7"
author: "YOUR NAME HERE"
date: "2023-10-18"
format: 
  html:
    embed-resources: true
    code-tools: true
    code-summary: "Code"
---

Remember, you must submit *both* your .Rmd and the compiled .html in order to receive full credit! In addition, to receive full credit, your code output and plots must be correctly formatted.

### Collaborators

```{r}
# LOAD ANY RELEVANT PACKAGES HERE
```

## A. Skewness and Kurtosis

Skewness describes how asymmetrical the distribution of a numerical variable is around its mean. Given observations $x_1,\ldots, x_n$, we can calculate the sample skewness $s$ of a variable using the following formula:

$$s = \frac{\frac{1}{n}\sum\limits_{i=1}^n(x_i-\overline{x})^3}{\left[\frac{1}{n}\sum\limits_{i=1}^n(x_i-\overline{x})^2\right]^{3/2}}$$
Kurtosis is a measure of the "tailedness" of the distribution of a numerical variable is around its mean. Higher values of kurtosis indicate more extreme outliers. Given observations $x_1,\ldots, x_n$, we can calculate the sample kurtosis $k$ of a variable using the following formula:

$$k = \frac{\frac{1}{n}\sum\limits_{i=1}^n(x_i-\overline{x})^4}{\left[\frac{1}{n}\sum\limits_{i=1}^n(x_i-\overline{x})^2\right]^{2}}-3$$

1. Write a function `skewness()` that takes as input a numeric vector `x` and returns the sample skewness. There are functions in R that compute skewness, but you cannot use any of them--write your own implementation.

2. Write a function `kurtosis()` that takes as input a numeric vector `x` and returns the sample skewness. There are functions in R that compute kurtosis, but you cannot use any of them--write your own implementation.

3. Use your function to compute the sample skewness of the `arr_delay` variable in the `flights` dataset contained in the `nycflights13` package.

4. Use your function to compute the sample kurtosis of the `arr_delay` variable in the `flights` dataset contained in the `nycflights13` package.

## B. Finding an error

Suppose you have two teams of runners participating in a 5k. We wish to write a function that takes as input two vectors representing the times of the runners in each team and returns a list of two vectors representing the ranks of each team's runners.

For example, if the first team's times are `c(16.8, 21.2, 19.1)` and the second team's times are `c(17.2, 18.1, 20.0)`, the function should return `c(1, 6, 4)` for the first team and `c(2, 3, 5)` for the second team. 

Below is a draft version of the function `get_runner_ranks()`. However, there is an error somewhere. Use any method we discussed in class to identify the error.

```{r, error = T, echo = T, eval = T}
get_runner_ranks <- function(x, y) {
  # combine all runner times
  combined_times <- c(x, y) 
  
  # sort all runner times from fastest to slowest
  sort(combined_times, decreasing = T)
  
  # create ranks vectors
  ranks_x <- numeric(length(x))
  ranks_y <- numeric(length(y))
  
  for (i in seq_along(ranks_x)) {
    # look up rank of time i in x in combined_times
    ranks_x[i] <- match(x[i], combined_times)
  }
  
  for (i in seq_along(ranks_y)) {
    # look up rank of time i in y in combined_times
    ranks_y[i] <- match(y[i], combined_times)
  }
  
  # return a list of first team and second team ranks
  return(list(x = ranks_x, y = ranks_y))
}
```

5. Below, write a corrected version of `get_runner_ranks()` and compute `get_runner_ranks(c(16.8, 21.2, 19.1), c(17.2, 18.1, 20.0))`.

```{r, error = T, echo = T, eval = T}
get_runner_ranks <- function(x, y) {
  # YOUR CODE HERE
}
```

6. Explain in your own words what the error was.
